---
title: "Project0 - Analysis of AIWIP GOWN Diversity and Denitrification potential"
author: "Gamaliel Lysander B. Cabria"
date: "08/11/2024"
output: html_notebook
---

# Project 0

## Analysis of the Amplicon Sequence Diversity


### DADA2 Pipeline
The input will be using raw reads that were not trimmed or undergone any quality control.

Core numbers can be assigned so that parallelization for filtering and chimera removal can be done. The default is 12.

```{r}
library(dada2)
library(tidyverse)

input<-"/home/glbcabria/Workbench/P0/qc" #Contains raw *fastq.gz 

n_cores <- 12
# Taking in inputs for SLURM
args <- commandArgs(trailingOnly = TRUE)
if (length(args) < 1) {
  n_cores <- 12
  cat("No number of cores provided. Defaulting to 12 cores.\n")
} else {
  n_cores <- as.numeric(args[1])
  if (is.na(n_cores) || n_cores < 1) {
    stop("Invalid number of cores.")
  }
}


# Listing inputs
forward_fq <- sort(list.files(input, pattern="_R1_paired.trim.fastq.gz", full.names = TRUE))
reverse_fq <- sort(list.files(input, pattern="_R2_paired.trim.fastq.gz", full.names = TRUE))
sample_names <- sapply(strsplit(basename(forward_fq), "_"), `[`, 1)

list_reads <- c(forward_fq, reverse_fq)

for (i in seq_along(forward_fq)){
    out_name <- paste0(input,'/',sample_names[i], ".qualplot.png")
    png(out_name, width=800, height = 1000)
    print(plotQualityProfile(c(forward_fq[i],reverse_fq[i])))
    dev.off()
}
```
Estimation of the size for use in `filtntrim` can be done using the red curves in this plot. I saved it alongside the input and per Sample Name.

```{r}
# Filter and Trimming
filt_for_fq <- file.path(input, "filtered", paste0(sample_names, "_F_filt.fastq.gz"))
filt_rev_fq <- file.path(input, "filtered", paste0(sample_names, "_R_filt.fastq.gz"))
names(filt_for_fq) <- sample_names
names(filt_rev_fq) <- sample_names

filtntrim <- filterAndTrim(
    forward_fq, filt_for_fq,
    reverse_fq, filt_rev_fq,
    truncLen = c(240,240),
    #maxReads = 100000, #Comment out on first run to see the distribution of reads
    maxN=0, maxEE=c(2,2), truncQ=2, 
    rm.phix=TRUE,
    compress=TRUE, 
    multithread=n_cores
)

# Error Rate
forward_err <- learnErrors(filt_for_fq, multithread = TRUE)
reverse_err <- learnErrors(filt_rev_fq, multithread = TRUE)

plot_errs_for <- plotErrors(forward_err, nominalQ = TRUE)
plot_errs_rev <- plotErrors(reverse_err, nominalQ = TRUE)
ggsave(plot = plot_errs_for, filename = paste0(input,'/plot_errs_for.png'))
ggsave(plot = plot_errs_rev, filename = paste0(input,'/plot_errs_rev.png'))
```
The errors are used for error-correction per sample. Similar to error correction in spades? Also visualization of Errors can sho what is the most common nucleotide shift.

```{r}
# Sample Inference
dada_for <- dada(filt_for_fq, forward_err, multithread = n_cores, pool = TRUE, DETECT_SINGLETONS=TRUE)
dada_rev <- dada(filt_rev_fq, reverse_err, multithread = n_cores, pool = TRUE, DETECT_SINGLETONS=TRUE)

dada_for[[1]]
denoisedF<-sapply(dada_for, function(x) sum(getUniques(x)))
denoisedR<-sapply(dada_rev, function(x) sum(getUniques(x)))
denoisedF
denoisedR

```
Inspecting the dada-class object can show how many sequence variants from number of input sequences

```{r}
merged_reads <- mergePairs(dada_for, filt_for_fq, dada_rev, filt_rev_fq, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(merged_reads[[1]])
```

```{r}
# Constructing sequence table
seqtab <- makeSequenceTable(merged_reads)
dim(seqtab)
merged_counts <- sapply(merged_reads, function(x) sum(x$abundance))
print(merged_counts)
# table(nchar(getSequences(seqtab)))
```

```{r}
# Remove Chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)

nochim_counts<- rowSums(seqtab.nochim)
nochim_counts
```

Tracking is made to check how much is the lost of each step

```{r}
# Tracking
getN <- function(x) sum(getUniques(x))
track <- cbind(filtntrim, sapply(dada_for, getN), sapply(dada_rev, getN), sapply(merged_reads, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample_names

final_track <- track %>% as.data.frame %>% 
mutate(Percent_Merged = merged / filtered ) %>% 
mutate(Percent_NoChimeras = nonchim / filtered )

write.csv(final_track, file = 'track.csv')
```

Output can be save as RData or csv files. Here I am saving it as CSV files that can be inputted in microeco or phyloseq object with minimal wrangling.

```{r}
### REMOVE THIS AFTER TESTRUN: THIS SECTION HAS NOT BEEN CHECKED ###

taxa <- assignTaxonomy(seqtab.nochim,
        "/work/ebg_lab/eb/AIWIP_GOWN_amplicons/DADA2/assignedTaxa/silva_nr99_v138.2_toGenus_trainset.fa.gz",
        multithread = TRUE, tryRC=TRUE)

taxa <- addSpecies(taxa,
        "/work/ebg_lab/eb/AIWIP_GOWN_amplicons/DADA2/assignedTaxa/silva_v138.2_assignSpecies.fa.gz")

taxa_df <- as.data.frame(taxa)

write.csv(taxa_df, file = 'taxa.csv', row.names = TRUE)

# Save OTU/ASV Table
otu_table <- as.data.frame(seqtab.nochim)
rownames(otu_table) <- sample_names
write.csv(otu_table, file = "otu_table.csv", row.names = TRUE)
```


### Analysis of Cluny and Canmore from 2019 and 2022 sampling season.
Per request of the Banfield Lab, we need to check the presence of ANME archaeon *Nitroperedens*

This uses two or three inputs: Taxa Table, OTU Table, and a Metadata Table. Taxa Table and OTU Table is directly from the output of DADA2. The Metadata table should be made separately.

```{r}
library(microeco)
library(data.table)
library(tidyverse)
library(janitor)

# Declaring Inputs
input_dir <- '/home/glbcabria/Workbench/gownproject/P0 - GOWN'
setwd(input_dir)

input_taxa <- fread(paste0(input_dir,'/taxa.csv'), sep = ',', header = TRUE, nThread = 16 ) 
input_otu <- fread(paste0(input_dir,'/otu_table.csv'), header = TRUE, nThread = 16) 
unique_ASV <- unique(c(input_taxa$V1, input_otu$V1))

ASV_df <- data.frame(
    ASV = unique_ASV) %>%
    mutate(
        ASV_ID = paste0("ASV_",row_number())
    )

tax_table <- input_taxa %>%
    left_join(ASV_df, by = c('V1' = 'ASV')) %>%
    column_to_rownames(var = "ASV_ID") %>%
    select(-V1)

otu_table <- input_otu %>%
    t() %>%
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    rownames_to_column('ASV') %>%
    left_join(ASV_df, by = c("ASV")) %>%
    select(-ASV) %>%
    column_to_rownames(var = "ASV_ID") %>%
    mutate(across(where(is.character), as.numeric))

metadata_files <- list.files(input_dir, pattern = ".*metadata.*\\.csv$", full.names = TRUE, ignore.case = TRUE)
if (length(metadata_files) > 0) {
# Read all matching files and bind them together
    input_metadata <- map_dfr(metadata_files, read_csv)
 } else {
# Empty tibble with no columns
    input_metadata <- tibble(
        samples = colnames(otu_table)
        )
 }
```

Metadata should be prepared around here. All code above can be reutilised. For the specific samples from Cluny and Canmore Tourist Info, I will create a metadata from the input file names and other data provided by Muhe and Daemon.

This section can be substituted or removed if metadata are known. 
```{r}
metadata_table <- input_metadata %>%
    mutate(
        SeqDate = str_split_i(samples, "-", -1),
        SeqCode = str_split_i(samples, "-", -2),
        Sequencer = str_split_i(samples, "-", 1)
    ) %>%
    column_to_rownames(var = "samples")
```

Setting up the microeco object based on the inputs above. 

```{r}
meco <- microtable$new(
  otu_table   = otu_table,
  tax_table   = tax_table,
  sample_table = metadata_table
)
```

The above code chunks can be combined as a single R code for quick analysis using the server.

Let us just analyse the Cluny and Canmore using pairwise analysis of relative abundance.

```{r}
library(ggpubr)
library(tidyverse)
library(rstatix)

cluny_canmore <- otu_table %>%
    select(c(
        'MDI-bact-GOWN-C21-20200316', # 19GWC19053 760 Canmore Info
        'MDI-bact-GOWN-C9-20200316',  # 19GWC19026 218 Cluny 
        'DMO-GOWN-bact-2023-G10-20230321', # 22GWC22053 760 Canmore
        'DMO-GOWN-bact-2023-G15-20230321' # 22GWC22075 218 Cluny 85-2 Deep
    ))

metadata_cluny_canmore <- input_metadata %>%
    filter(
        samples %in% c(
        'MDI-bact-GOWN-C21-20200316', # 19GWC19053 760 Canmore Info
        'MDI-bact-GOWN-C9-20200316',  # 19GWC19026 218 Cluny 
        'DMO-GOWN-bact-2023-G10-20230321', # 22GWC22053 760 Canmore
        'DMO-GOWN-bact-2023-G15-20230321' # 22GWC22075 218 Cluny 85-2 Deep
    )
    ) %>%
    mutate(
        Season = ifelse(grepl("MDI", samples), 2022, 2019),
        Site = ifelse(grepl("C21|G10", samples), '760 Canmore', '218 Cluny 85-2 Deep')
    )

cluny_canmore_long <- cluny_canmore %>%
    rownames_to_column(var = 'ASV') %>%
    pivot_longer(-ASV, names_to = "samples", values_to = "count") %>%
    left_join(metadata_cluny_canmore, by = "samples")    

# Compute Alpha Diversity
alpha_metrics <- cluny_canmore_long %>%
  group_by(samples, Site, Season) %>%
  summarise(
    Observed = sum(count > 0),
    .groups = "drop_last"
  ) %>%
  left_join(
    cluny_canmore_long %>%
      group_by(samples, Site, Season) %>%
      mutate(p = count / sum(count, na.rm = TRUE)) %>%
      summarise(
        Shannon = -sum(if_else(p > 0, p * log(p), 0), na.rm = TRUE),
        Simpson = 1 - sum(p^2, na.rm = TRUE),
        InvSimpson = 1 / sum(p^2, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("samples","Site","Season")
  ) %>%
  mutate(Evenness = Shannon / log(Observed)) %>%
  ungroup() %>%
  pivot_longer(cols = c(Observed, Shannon, Simpson, InvSimpson, Evenness),
               names_to = "Metric", values_to = "Value") %>%
  mutate(Season = factor(Season))


# Stats
stats <- alpha_metrics %>%
  group_by(Site, Metric) %>%
  wilcox_test(Value ~ Season, paired = TRUE) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  ungroup()

# --- 6) Plot
plot_alpha <- ggplot(alpha_metrics, aes(x = factor(Season), y = Value, color = Site, group = Site)) +
  geom_point(size = 3) +
  geom_line(aes(group = Site), linewidth = 0.8) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(x = "Season", y = "Alpha Diversity", color = "Site",
       title = "Alpha diversity comparison between 2019 and 2023") +
  theme_minimal(base_size = 14) +
  theme(
        panel.background = element_rect(fill = "white", colour = NA),
    plot.background  = element_rect(fill = "white", colour = NA),
    strip.background = element_rect(fill = "white", colour = NA)
  )

# Add significance labels
plot_alpha + stat_pvalue_manual(
  stats,
  label = "p.adj.signif",
  y.position = alpha_metrics %>% group_by(Metric) %>% summarise(max_val = max(Value)*1.1) %>% pull(max_val),
  tip.length = 0.01
)

plot_alpha
```

Alpha diversity is not to show there is no significant change between the sites and time point. It lack enough replicates.

Here, let us do a simpled paired-test per ASV
```{r}
# Make relative abundances (optional, but recommended)
asv_rel <- cluny_canmore_long %>%
  group_by(samples) %>%
  mutate(rel_abund = count / sum(count, na.rm = TRUE)) %>%
  ungroup()

# Pairwise test for each ASV
asv_stats <- asv_rel %>%
  group_by(ASV, Site) %>%
  wilcox_test(rel_abund ~ Season, paired = TRUE) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  ungroup()

# 2) Plot: connect each ASV across Seasons, per Site
plot_asv_rel_change <- ggplot(
  asv_rel,
  aes(x = Season, y = rel_abund, group = ASV)
) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1.7) +
  facet_wrap(~ Site, scales = "free_y") +
  labs(
    x = NULL, y = "Relative abundance",
    title = "ASV changes between seasons (per Site)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    aspect.ratio = 1.2,
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background  = element_rect(fill = "white", colour = NA),
    strip.background = element_rect(fill = "white", colour = NA)
  )

plot_asv_rel_change

# ggsave(filename = "/home/glbcabria/Desktop/plot_asv_rel_change.png", plot = plot_asv_rel_change, 
#   units = c('px'), dpi = 150, height = 1500, width = 1500
# )

# ggsave(filename = "/home/glbcabria/Desktop/plot_cluny_canmore_alpha.png", plot = plot_alpha,
#   units = c('px'), dpi = 150, height = 1500, width = 1500
# )

```

