---
title: "Project 0 - Identification of Microbial Effect in GW Denitrification Project"
output: html_notebook
---

## Intro
This is a notebook for the bioinformatic analysis/workflow for identifying the effect of microbes in the groundwater (GW) wells **in situ**. The sections will be based on the analysis that has been done and will be added as I go. First, is selecting wells for sampling

## Field sampling 2025
Importing the files for analysis
```{r}
library(tidyverse)
library(readxl)
library(janitor)

clean_name <- function(names) {
  names <- gsub("\\s+", "_", names)  # Replace multiple whitespaces with a single underscore
  make.unique(names, sep = "_")      # Ensure names are still unique
}

input_collectionteam20255<-'/home/glbcabria/Workbench/GOWNData/EPA 2025 GW_Quality_Well List 2025 U of C.xlsx'
input_darkoxygenpsupp<-'/home/glbcabria/Workbench/GOWNData/EmilsSupp_41467_2023_38523_MOESM4_ESM.xlsx'
input_2023GOWNMetadata <-'/home/glbcabria/Workbench/GOWNData/2023 Dec GOWN Metadata Stn Public.xls'
input_denitrificationpotential <-'/home/glbcabria/Workbench/GOWNData/denitrification potential.xlsx'

collectionteam2025<-read_excel(input_collectionteam20255, sheet = 'GW_Quality_well list 2025-26', 
                               col_names = TRUE, .name_repair = clean_name )
darko2supp<-read_excel(input_darkoxygenpsupp, sheet = 'Chem_Phys_Gas_Isotope_Microbiol', skip=1, 
                       col_names = TRUE, .name_repair = clean_name)
GOWNMetadata<-read_excel(input_2023GOWNMetadata, sheet = 'Detail', col_names = TRUE, 
                         .name_repair = clean_name)
deNO3potential<-read_excel(input_denitrificationpotential, sheet = 'denitrifiers by sample', col_names = TRUE,
                           .name_repair = clean_name, n_max = 26)
```

Let us filter only the needed rows and columns per input so we can have clean merge of dataframes later on.
```{r}
clean_dnp <- deNO3potential %>%
  clean_names() %>%
  separate(x14, into=c("GOWN_ID","GOWN_NAME_DNP"), sep = ' ', extra='merge') %>%
  select(!starts_with("x")) %>%
  select(c('GOWN_ID','GOWN_NAME_DNP','denitrifier_relative_abundance_percent','nitrate','index_percent_d_percent_o',)) %>%
  mutate(index_percent_d_percent_o=gsub('\\?',NA, index_percent_d_percent_o))

clean_metadata <- GOWNMetadata %>%
  select(Station_Code_Gown,STATION_NAME,AWWID_ID,Latitude,Longitude,Completion,Formation,Lithology,Aquifer,Depth_Class) %>%
  mutate(Station_Code_Gown=as.character(Station_Code_Gown))

clean_darkoxsupp <- darko2supp %>%
  select(GOWN_Well_Number, GOWN_Well_Name,
         Chem_METHANE_ugL,Chem_ETHANE_ugL,Chem_PROPANE_ugL,
         Chem_CO2_mgL,Chem_O2_mgL,Chem_O2_FIELD_METER_mgL,
         Chem_Cl_dis_mgL,Chem_SO4_dis_mgL,Chem_Bicarbonate_HCO3_calc_mgL,
         Chem_Fe_dis_mgL,Chem_Mn_dis_mgL,
         Chem_NH3_mgL,Chem_N_NO3_mgL,Chem_N_NO2_mgL
         )

clean_samplingtarget <- collectionteam2025 %>%
  clean_names() %>%
  filter(gown_no != "SUBTOTAL" & !is.na(gown_no) ) %>%
  mutate(COLLECTIONWELL_NAME = name) %>%
  select(gown_no, COLLECTIONWELL_NAME, wds_station_no, office) %>%
  separate_rows(gown_no, sep = ',')
  
join_df <- clean_samplingtarget %>%
  left_join(clean_metadata, by = c("gown_no"="Station_Code_Gown") ) %>%
  left_join(clean_dnp, by = c("gown_no" = "GOWN_ID") ) %>%
  left_join(clean_darkoxsupp, by = c("gown_no" = "GOWN_Well_Number") )

join_dnp_metadata <- clean_dnp %>%
  left_join(clean_metadata, by = c("GOWN_ID" = "Station_Code_Gown" )) %>%
  left_join(clean_darkoxsupp, by = c("GOWN_ID" = "GOWN_Well_Number") ) %>%
  left_join(clean_samplingtarget %>% select(gown_no,office), by=c("GOWN_ID"="gown_no"))
  
```


